---
title: "LPS853 CRISPR screen: drug vs DMSO analysis"
output: html_document: true
   
---

```{r setup, include=FALSE}
# Define paths
root_dir <- rprojroot::find_root(rprojroot::has_dir("LPS853_Brunello"))
analysis_dir <- file.path(root_dir, "LPS853_Brunello")
plots_dir <- file.path(root_dir, "LPS853_Brunello", "Plot_MAGecK")

# Create directories if they don't exist
if (!dir.exists(plots_dir)) {dir.create(plots_dir, recursive = TRUE)}
if (!dir.exists(analysis_dir)) {dir.create(analysis_dir, recursive = TRUE)}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
```

# Overview

The data are read counts from a Brunello genome-wide screen in LPS853 cells:
* about 19,000 genes with 4 sgRNAs each, plus 1,000 non-targeting guides
* one sample per condition: T0, DMSO, Palbociclib, Doxorubicin and Nutlin-3a (no replicates)

# 0. Packages and parameters

```{r packages}
# install.packages(c("readxl","MASS","dplyr","ggplot2","ggrepel","patchwork","pROC","openxlsx"))
library(readxl)
library(MASS)       # rlm() for the robust growth-correction fit
library(dplyr)
library(ggplot2)
library(ggrepel)
library(patchwork)
library(pROC)
library(openxlsx)

xlsx_file   <- file.path(root_dir,"can-25-1858_supplementary_table_s4_suppst4.xlsx")  # path to excel sheet readcounts
min_T0      <- 30     # drop sgRNAs with fewer than this many reads at T0
sd_window   <- 1001   # sgRNAs per sliding window for abundance-matched SD
fdr_cut     <- 0.1
set.seed(1)

col_sens <- "#2a6fdb"; col_res <- "#d9661f"; col_ns <- "grey75"
theme_set(theme_classic(base_size = 10))
```

# 1. Load data

The count sheet has a title row, so the header is on row 2 (`skip = 1`).

```{r load}
counts <- read_excel(xlsx_file, sheet = "Readcounts CRISPR screens", skip = 1)
counts <- as.data.frame(counts)
samples <- c("LPS853_T0", "LPS853_Ctrl", "LPS853_Palbo", "LPS853_Dox", "LPS853_Nut")
counts[samples] <- lapply(counts[samples], as.numeric)
dim(counts)
head(counts)

# Paper MAGeCK gene scores (for comparison)
mageck <- as.data.frame(read_excel(xlsx_file, sheet = "MAGeCK output"))

# Reference essential / non-essential genes (Hart et al.)
ref <- read_excel(xlsx_file, sheet = "(Non-) essential gene lists")
essential    <- na.omit(ref[["Essential genes"]])
nonessential <- na.omit(ref[["Non-essential genes"]])
c(essential = length(essential), nonessential = length(nonessential))

is_nt <- counts$gene == "NonTargeting"
table(is_nt)
```

# 2. Library-level QC and normalisation

## 2.1 Raw count summary

```{r rawqc}
gini <- function(x) {
  x <- sort(x); n <- length(x)
  sum((2 * seq_len(n) - n - 1) * x) / (n * sum(x))
}
qc_tab <- data.frame(
  sample        = samples,
  total_reads   = sapply(counts[samples], sum),
  median_reads  = sapply(counts[samples], median),
  zero_sgRNAs   = sapply(counts[samples], function(x) sum(x == 0)),
  gini          = round(sapply(counts[samples], gini), 3)
)
qc_tab
```

A Gini index around 0.1–0.2 at T0 is ideal and anything below about 0.35 is acceptable. A rise
after selection is expected, because essential-gene guides drop out.

## 2.2 Median-ratio normalisation

```{r norm}
cnt <- as.matrix(counts[samples])
lg  <- log(cnt); lg[!is.finite(lg)] <- NA          # zero counts are excluded from the geometric mean
geo <- rowMeans(lg)                                # per-sgRNA log geometric mean
size_factors <- exp(apply(lg - geo, 2, median, na.rm = TRUE))
round(size_factors, 3)

norm <- sweep(cnt, 2, size_factors, "/")
L    <- log2(norm + 1)                             # log2 normalised counts
```

## 2.3 Sample correlation

```{r corr}
round(cor(L), 3)
```

The drug arms correlate with DMSO at r ≈ 0.985, so the drug effects are subtle. All
four selected samples correlate less with T0 (r ≈ 0.84–0.87).

## 2.4 Essential-gene dropout: ROC / AUC and NNMD

If the screen worked, essential genes should be depleted relative to non-essential genes in every
arm compared with T0. For each arm we compute gene-level mean log2FC vs T0 (centred on non-targeting
guides), then the ROC AUC and NNMD. NNMD is the difference between essential and non-essential means,
divided by the non-essential SD; values below −2 are good and below −3 are excellent.

```{r auc}
gene_lfc_vs_T0 <- function(s) {
  lfc <- L[, s] - L[, "LPS853_T0"]
  lfc <- lfc - median(lfc[is_nt])
  tapply(lfc, counts$gene, mean)
}

ess_qc <- lapply(samples[-1], function(s) {
  g  <- gene_lfc_vs_T0(s)
  ge <- g[names(g) %in% essential]
  gn <- g[names(g) %in% nonessential]
  r  <- roc(response  = c(rep(1, length(ge)), rep(0, length(gn))),
            predictor = c(ge, gn), direction = ">", quiet = TRUE)   # lower LFC = essential
  list(sample = s, roc = r, ge = ge, gn = gn,
       stats = data.frame(sample = s,
                          AUC = round(as.numeric(auc(r)), 3),
                          NNMD = round((mean(ge) - mean(gn)) / sd(gn), 2),
                          mean_LFC_essential = round(mean(ge), 2),
                          mean_LFC_nonessential = round(mean(gn), 2)))
})
names(ess_qc) <- samples[-1]
do.call(rbind, lapply(ess_qc, `[[`, "stats"))
```

```{r roc-plot, fig.width = 10, fig.height = 4.2}
# Histogram for DMSO vs T0
dmso <- ess_qc[["LPS853_Ctrl"]]
hist_df <- rbind(data.frame(LFC = dmso$gn, set = "Non-essential"),
                 data.frame(LFC = dmso$ge, set = "Essential"))
p_hist <- ggplot(hist_df, aes(LFC, fill = set)) +
  geom_histogram(bins = 60, position = "identity", alpha = 0.85) +
  scale_fill_manual(values = c("Essential" = col_sens, "Non-essential" = "grey65")) +
  labs(title = "DMSO vs T0", x = "Gene mean log2FC", y = "Genes", fill = NULL) +
  theme(legend.position = c(0.25, 0.85))

# ROC curves for all arms
roc_df <- do.call(rbind, lapply(ess_qc, function(z) {
  data.frame(sample = sub("LPS853_", "", z$sample),
             FPR = 1 - z$roc$specificities, TPR = z$roc$sensitivities)
}))
roc_df$sample <- factor(roc_df$sample, levels = c("Ctrl", "Palbo", "Dox", "Nut"),
                        labels = c("DMSO", "Palbo", "Dox", "Nutlin"))
auc_lab <- sapply(ess_qc, function(z) z$stats$AUC)
p_roc <- ggplot(roc_df, aes(FPR, TPR, colour = sample)) +
  geom_abline(linetype = 2, colour = "grey60") +
  geom_path(linewidth = 0.7) +
  scale_colour_manual(values = c("#1b1b1b", "#2a6fdb", "#d9661f", "#7a3fb0"),
                      labels = paste0(c("DMSO", "Palbo", "Dox", "Nutlin"),
                                      " (AUC ", auc_lab, ")")) +
  coord_equal() +
  labs(title = "Essential vs non-essential ROC (vs T0)",
       x = "False positive rate (non-essential)", y = "True positive rate (essential)",
       colour = NULL) +
  theme(legend.position = c(0.65, 0.2))

p_hist + p_roc
```

The essential-gene mean LFC is weaker in the drug arms (about −1.4) than in DMSO (−1.73). That
means the drug-treated cells went through fewer doublings. This matters for step 4.

# 3. Helper: abundance-matched guide Z and gene score

Low-count guides are noisier. Each guide's LFC is therefore divided by the SD of guides with
similar DMSO abundance, using a sliding window of `sd_window` guides sorted by DMSO log count.
Gene Z is `sum(guide z) / sqrt(n guides)`, the same approach as drugZ. The growth correction
uses a robust (bisquare) regression line fitted on all guides.

```{r helpers}
# Sliding-window SD (centred; window shrinks at the edges), sample SD (n-1)
rolling_sd <- function(x, w) {
  n <- length(x); h <- (w - 1) %/% 2
  cs  <- c(0, cumsum(x)); cs2 <- c(0, cumsum(x^2))
  lo  <- pmax(1, seq_len(n) - h); hi <- pmin(n, seq_len(n) + h)
  k   <- hi - lo + 1
  s1  <- cs[hi + 1] - cs[lo]; s2 <- cs2[hi + 1] - cs2[lo]
  sqrt((s2 - s1^2 / k) / (k - 1))
}

gene_z <- function(lfc, abundance, gene, w = sd_window) {
  o  <- order(abundance)
  sd <- numeric(length(lfc)); sd[o] <- rolling_sd(lfc[o], w)
  z  <- lfc / sd
  data.frame(gene = gene, z = z, lfc = lfc) |>
    group_by(gene) |>
    summarise(n_sgRNA = n(), LFC = mean(lfc), Z = sum(z) / sqrt(n()), .groups = "drop")
}

# Robust straight-line fit (Tukey bisquare M-estimator) on ALL guides.
# Guides with real drug effects are outliers and get down-weighted, so the line
# describes the bulk "same biology, fewer doublings" relationship. Deterministic.
robust_fit <- function(x, y) {
  f <- rlm(y ~ x, psi = psi.bisquare, maxit = 100)
  c(slope = unname(coef(f)[2]), intercept = unname(coef(f)[1]))
}
```

# 4. Drug vs DMSO

## 4.1 Filter and set up log fold changes

```{r filter}
keep  <- counts$LPS853_T0 >= min_T0
gene  <- counts$gene[keep]
nt_k  <- is_nt[keep]
Lk    <- L[keep, ]
sum(keep); sum(!keep)

centre <- function(v) v - median(v[nt_k])
lfc_dmso_T0 <- centre(Lk[, "LPS853_Ctrl"] - Lk[, "LPS853_T0"])   # baseline fitness effect
abund_dmso  <- Lk[, "LPS853_Ctrl"]
```

## 4.2 Why a growth correction is needed

Plotting drug-vs-T0 against DMSO-vs-T0 for every guide shows a slope below 1. Essential-gene guides
dropped out less in the drug arms because those cells divided fewer times. A naive "drug minus DMSO"
comparison therefore makes essential genes look like *resistance* hits. This explains why
GINS2, CIAO1 and ribosomal genes top the paper's resistance lists.

```{r growth, fig.width = 10, fig.height = 3.6}
arms <- c(Palbo = "LPS853_Palbo", Dox = "LPS853_Dox", Nutlin = "LPS853_Nut")
paper_col <- c(Palbo = "Score_palbo", Dox = "Score_dox", Nutlin = "Score_nutlin")

fits <- lapply(arms, function(s) {
  y   <- centre(Lk[, s] - Lk[, "LPS853_T0"])
  list(y = y, fit = robust_fit(lfc_dmso_T0, y))
})
sapply(fits, function(f) round(f$fit, 3))

gp <- lapply(names(fits), function(k) {
  f  <- fits[[k]]
  df <- data.frame(x = lfc_dmso_T0, y = f$y)[sample(length(f$y), 15000), ]
  ggplot(df, aes(x, y)) +
    geom_point(size = 0.3, alpha = 0.3) +
    geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey50") +
    geom_abline(slope = f$fit["slope"], intercept = f$fit["intercept"],
                colour = col_res, linewidth = 0.8) +
    labs(title = sprintf("%s (slope %.2f)", k, f$fit["slope"]),
         x = "DMSO vs T0 log2FC", y = paste(k, "vs T0 log2FC"))
})
wrap_plots(gp, nrow = 1)
```

## 4.3 Gene scores

For each drug, this chunk computes:

* `LFC_vs_DMSO`, `Z_raw`: direct drug-vs-DMSO comparison, comparable to the paper's MAGeCK score
* `LFC_adj`, `Z_adj`: the growth-corrected residual (**the main result**)
* `p_sens` / `FDR_sens`: depleted in drug, meaning knockout sensitises
* `p_res` / `FDR_res`: enriched in drug, meaning knockout gives resistance

```{r scores}
dmso_T0_gene <- tapply(lfc_dmso_T0, gene, mean)

results <- lapply(names(arms), function(k) {
  s   <- arms[[k]]
  raw <- centre(Lk[, s] - Lk[, "LPS853_Ctrl"])
  adj <- fits[[k]]$y - (fits[[k]]$fit["slope"] * lfc_dmso_T0 + fits[[k]]$fit["intercept"])

  R <- gene_z(raw, abund_dmso, gene)
  A <- gene_z(adj, abund_dmso, gene)

  out <- data.frame(gene = A$gene, n_sgRNA = A$n_sgRNA,
                    DMSO_vs_T0_LFC = as.numeric(dmso_T0_gene[A$gene]),
                    LFC_vs_DMSO = R$LFC[match(A$gene, R$gene)],
                    Z_raw       = R$Z[match(A$gene, R$gene)],
                    LFC_adj = A$LFC, Z_adj = A$Z) |>
    filter(gene != "NonTargeting") |>
    mutate(p_sens   = pnorm(Z_adj),
           p_res    = pnorm(Z_adj, lower.tail = FALSE),
           FDR_sens = p.adjust(p_sens, "BH"),
           FDR_res  = p.adjust(p_res,  "BH"),
           paper_MAGeCK_score = mageck[[paper_col[[k]]]][match(gene, mageck$Gene)],
           likely_artefact_essential = DMSO_vs_T0_LFC < -1) |>
    arrange(Z_adj)
  out
})
names(results) <- names(arms)

# Hit counts and agreement with the paper's MAGeCK scores
data.frame(
  arm        = names(results),
  sens_FDR10 = sapply(results, function(r) sum(r$FDR_sens < fdr_cut)),
  res_FDR10  = sapply(results, function(r) sum(r$FDR_res  < fdr_cut)),
  rho_raw_vs_paper = sapply(results, function(r)
    round(cor(r$Z_raw, r$paper_MAGeCK_score, method = "spearman", use = "complete"), 3)),
  rho_adj_vs_paper = sapply(results, function(r)
    round(cor(r$Z_adj, r$paper_MAGeCK_score, method = "spearman", use = "complete"), 3))
)
```

## 4.4 Top hits per arm

```{r tophits}
show_cols <- c("gene", "DMSO_vs_T0_LFC", "LFC_vs_DMSO", "LFC_adj", "Z_adj",
               "FDR_sens", "FDR_res", "paper_MAGeCK_score")
for (k in names(results)) {
  cat("\n###", k, "— top sensitising\n")
  print(head(results[[k]][show_cols], 15), digits = 3, row.names = FALSE)
  cat("\n###", k, "— top resistance\n")
  print(head(results[[k]][order(-results[[k]]$Z_adj), show_cols], 10), digits = 3, row.names = FALSE)
}
```

Genes with `DMSO_vs_T0_LFC < -1` are strong essentials with few reads left. When they appear as
hits (for example COPA, DNAJC17 and PRKRA, which show up in all three arms), treat them with caution.

```{r shared}
Reduce(intersect, lapply(results, function(r) r$gene[r$FDR_sens < fdr_cut]))
```

## 4.5 How much do hits depend on the growth-correction slope?

The slope is estimated, not known. Re-scoring at slope ±0.03 shows which hits are robust.
The number of *sensitising* hits is very sensitive to the slope: a slightly lower slope pushes
many essential genes past the threshold. The strongest hits stay put, though. Trust genes that
remain significant across the whole range more than the exact hit counts.

```{r slope-sensitivity}
score_at_slope <- function(k, b) {
  y   <- fits[[k]]$y
  adj <- y - (b * lfc_dmso_T0 + fits[[k]]$fit["intercept"])   # same form as section 4.3
  A   <- gene_z(adj, abund_dmso, gene) |> filter(gene != "NonTargeting")
  A$FDR_sens <- p.adjust(pnorm(A$Z), "BH")
  A$FDR_res  <- p.adjust(pnorm(A$Z, lower.tail = FALSE), "BH")
  A
}
sens_tab <- do.call(rbind, lapply(names(arms), function(k) {
  b0 <- fits[[k]]$fit["slope"]
  do.call(rbind, lapply(c(-0.03, 0, 0.03), function(d) {
    A <- score_at_slope(k, b0 + d)
    data.frame(arm = k, slope = round(b0 + d, 3),
               sens_FDR10 = sum(A$FDR_sens < fdr_cut), res_FDR10 = sum(A$FDR_res < fdr_cut))
  }))
}))
sens_tab

# Sensitising hits found at all three slopes
robust_hits <- lapply(names(arms), function(k) {
  b0 <- fits[[k]]$fit["slope"]
  Reduce(intersect, lapply(c(-0.03, 0, 0.03), function(d) {
    A <- score_at_slope(k, b0 + d); A$gene[A$FDR_sens < fdr_cut] }))
})
names(robust_hits) <- names(arms)
lengths(robust_hits)
for (k in names(robust_hits)) {
  r <- results[[k]]; r <- r[r$gene %in% robust_hits[[k]] & !r$likely_artefact_essential, ]
  cat(k, ": ", paste(head(r$gene, 20), collapse = ", "), "\n", sep = "")
}
```

# 5. Pathway check: NHEJ

The test uses Stouffer's method on gene Z_adj. A negative value means the whole set sensitises.

```{r nhej}
nhej <- c("PRKDC", "XRCC4", "XRCC5", "XRCC6", "LIG4", "NHEJ1", "DCLRE1C")
do.call(rbind, lapply(names(results), function(k) {
  z <- results[[k]]$Z_adj[match(nhej, results[[k]]$gene)]; z <- z[!is.na(z)]
  sz <- sum(z) / sqrt(length(z))
  data.frame(arm = k, n = length(z), mean_Z = round(mean(z), 2),
             set_Z = round(sz, 2), p_sensitising = signif(pnorm(sz), 3))
}))

panel <- c(nhej, "TDP2", "TP53BP1", "ATM", "ABCC1", "TOP2A", "TOP2B",
           "RB1", "CDK2", "CCNE1", "CDKN2A", "TP53", "CDKN1A", "MDM2", "USP28")
panel_tab <- sapply(results, function(r) round(r$Z_adj[match(panel, r$gene)], 2))
rownames(panel_tab) <- panel
panel_tab
```

```{r heat, fig.width = 4.5, fig.height = 6}
hm <- as.data.frame(as.table(panel_tab)); names(hm) <- c("gene", "arm", "Z")
hm$gene <- factor(hm$gene, levels = rev(panel))
lim <- 8
ggplot(hm, aes(arm, gene, fill = pmax(pmin(Z, lim), -lim))) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.1f", Z)), size = 2.8) +
  scale_fill_gradient2(low = col_sens, mid = "grey95", high = col_res,
                       limits = c(-lim, lim), name = "Z_adj\n(capped ±8)") +
  labs(x = NULL, y = NULL, title = "Selected genes (growth-corrected Z)")
```

# 6. Rank plots

```{r rankplots, fig.width = 13, fig.height = 4.5}
lab_genes <- list(
  Palbo  = c("RB1", "CDKN2A", "FZR1", "CDKN1A", "TP53", "CDK2", "CCNE1", "E2F3", "KEAP1", "CKS1B"),
  Dox    = c("TP53", "CDKN1A", "SMAD4", "PRKDC", "XRCC4", "TDP2", "ABCC1", "XRCC6", "NHEJ1", "LIG4"),
  Nutlin = c("TP53", "CDKN1A", "USP28", "TP53BP1", "ATM", "SLC33A1", "SBDS", "OBFC1"))

rank_plot <- function(k) {
  r <- results[[k]] |>
    mutate(rank = rank(Z_adj, ties.method = "first"),
           class = case_when(FDR_sens < fdr_cut ~ "Sensitising",
                             FDR_res  < fdr_cut ~ "Resistance",
                             TRUE ~ "NS"))
  ggplot(r, aes(rank, Z_adj)) +
    geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.3) +
    geom_point(data = filter(r, class == "NS"), colour = col_ns, size = 0.3) +
    geom_point(data = filter(r, class != "NS"), aes(colour = class), size = 1) +
    geom_text_repel(data = filter(r, gene %in% lab_genes[[k]]), aes(label = gene),
                    size = 2.8, min.segment.length = 0, segment.size = 0.3,
                    nudge_x = ifelse(filter(r, gene %in% lab_genes[[k]])$Z_adj > 0, -4000, 4000),
                    direction = "y", max.overlaps = Inf) +
    scale_colour_manual(values = c(Sensitising = col_sens, Resistance = col_res)) +
    annotate("text", x = 0, y = max(r$Z_adj), hjust = 0, vjust = 1, size = 3,
             label = sprintf("Sensitising FDR<%.1f: %d\nResistance FDR<%.1f: %d",
                             fdr_cut, sum(r$class == "Sensitising"),
                             fdr_cut, sum(r$class == "Resistance"))) +
    labs(title = paste(k, "vs DMSO (growth-corrected)"), x = "Gene rank", y = "Gene Z") +
    theme(legend.position = "none")
}
wrap_plots(lapply(names(results), rank_plot), nrow = 1)
```

Raw versus corrected scores for Dox show how the correction moves essential genes out of the
"resistance" side:

```{r rawvsadj, fig.width = 5.5, fig.height = 4.5}
r <- results$Dox
ggplot(r, aes(Z_raw, Z_adj, colour = DMSO_vs_T0_LFC < -1)) +
  geom_point(size = 0.4, alpha = 0.5) +
  geom_abline(linetype = 2, colour = "grey50") +
  scale_colour_manual(values = c(`FALSE` = "grey60", `TRUE` = col_sens),
                      labels = c("Other", "Strong essential (DMSO vs T0 < -1)"), name = NULL) +
  labs(title = "Dox: raw vs growth-corrected gene Z", x = "Z raw (drug vs DMSO)", y = "Z adjusted") +
  theme(legend.position = "bottom")
```

# 7. Export

```{r export}
wb <- createWorkbook()
addWorksheet(wb, "QC")
writeData(wb, "QC", cbind(qc_tab, size_factor = round(size_factors, 3)))
writeData(wb, "QC", do.call(rbind, lapply(ess_qc, `[[`, "stats")), startRow = 9)
addWorksheet(wb, "Gene panel (Z_adj)")
writeData(wb, "Gene panel (Z_adj)", panel_tab, rowNames = TRUE)
for (k in names(results)) {
  sh <- paste(k, "vs DMSO"); addWorksheet(wb, sh)
  writeData(wb, sh, results[[k]]); freezePane(wb, sh, firstRow = TRUE, firstCol = TRUE)
}
saveWorkbook(wb, "LPS853_CRISPR_drug_vs_DMSO_results_R.xlsx", overwrite = TRUE)

ggsave("LPS853_rankplots.png", wrap_plots(lapply(names(results), rank_plot), nrow = 1),
       width = 13, height = 4.5, dpi = 200)
ggsave("LPS853_QC_ROC.png", p_hist + p_roc, width = 10, height = 4.2, dpi = 200)
```

# Session info

```{r session}
sessionInfo()
```
